Kepler's Demographic Cliff: Ages of Young Stars and Planets in the Kepler Field

### 2024-04-29 paper/ms.tex
update ms

% <250 words
\begin{abstract}
  Recent studies of stars in open clusters have shown how the rotation
  rates and lithium content of FGK stars evolve during the first few
  billion years of their lives.
  %
  Here we apply this knowledge to stars observed by the primary Kepler
  mission to curate catalogs of rotation and lithium-based ages.
  %
  Rotation periods are drawn from previous Kepler analyses; lithium is
  measured from new and archival HIRES spectra.
  %
  We report rotation-based ages for \nuniqstarsantosrotgyroappl\ stars
  and \nplwgyroagenograzing\ planets for which we believe
  gyrochronology is valid;
  these rotation-based ages are $\approx$15-20$\times$ more precise
  than isochrone ages for our sample of FGK stars.
  %
  While the independent lithium ages are consistent with XX\% of the rotation ages,
  this method is significantly less precise than rotation-based age dating
  for most of our sample.
  %
  Our resulting catalog of planetary ages includes
  \nplyounggyrotwosigma\ planets younger than 1\,Gyr at 2$\sigma$
  (\nplyounggyrothreesigma\ planets at 3$\sigma$).
  %
  In the broader sample of stars in the Kepler field, we find that the
  rotation-based age distribution decreases linearly from 3\,Gyr ago
  to present day, implying a linearly decreasing star formation rate.
  %
  While this parent stellar age distribution has implications for the
  star formation history of our Galaxy, it also bounds the relative
  balance of young and old planets detectable using both Kepler and
  also future wide-field transit surveys.
  %
\end{abstract}

@@

@@

\subsubsection{Systems with Rotation and Lithium in Disagreement}

@@

\startlongtable
\begin{deluxetable*}{llllllllrrrrrrl}
	%\tabletypesize{\footnotesize}
	\tabletypesize{\scriptsize}
  \tablecaption{Ages of Kepler planets and planet candidates.  This
  version of the table is truncated to include the youngest fifty
  confirmed Kepler planets, and the twenty youngest candidates,
  sorted by the minimum of either the rotation or lithium-based
  age.  The
  full machine-readable table contains ages and age limits for
  \nnonfopkoissomeageinfo\ non-false positive KOIs with MES$>$10.  
  \label{tab:planets}}
	%\toprule
	%\midrule
	%\endhead
	\startdata
  KOI & Kepler &  $T_{\rm eff}$ & $P_{\rm rot}$ & EW$_{\rm Li}$ &
  $t_{\rm gyro}$ & $t_{\rm Li} $ & Consistent? & $R_{\rm p}$ & $P_{\rm orb}$ &
  $Q_{\rm planet}$ & $Q_{\rm gyro}$ &  Spec? & Comment \\
  -- &   -- & K & days & m\AA & Myr & Myr &  str &   Earths &    days &       int  & int & bool & -- \\
	\hline
	\input{table_subgyr_confirmed} %\\
%	\rule
	\input{table_subgyr_candidate}
	\enddata
	\tablecomments{``Confirmed'' planets appear before``candidate'' planets.
		The bit quality flags for the rotation-based ages,
		$Q_{\rm gyro}$, are described in Section~\ref{subsec:flags}.
		Concisely summarized, they are:
		Bit 0: $T_{\rm eff}/{\rm K} \in [ 3800-6200]$ ?
		Bit 1: $\log g$$<$4.2?
		Bit 2: $M_{\rm G}$<3.9?
		Bit 3: In KEBC?
		Bit 4: Large $d_{\rm xm,Kep-Gaia}$?
		Bit 5: Confused Kep-Gaia crossmatch?
		Bit 6: Gaia DR3 non-single star?
		Bit 7: RUWE$>$1.4?
		Bit 8: Crowded?
		Bit 9: Far from main-sequence?
		Bit 10: \citetalias{Santos_2021} CP/CB?
		As an example, Kepler-1627 has $Q_{\rm gyro}$ flagged with bit 10 and bit 7.
		The analogous planet quality bit-flag, $Q_{\rm planet}$, has the
		following meaning.
		Bit 0: Candidate not reliable?
		Bit 1: MES$<$10?
		Bit 2: Grazing?
	}
\end{deluxetable*}

@@

\appendix
%\section{Can you avoid appendices?}
%\label{app:interp}

>>>>

% <250 words
\begin{abstract}
  Recent studies of stars in open clusters have shown how the rotation
  rates and lithium content of FGK stars evolve during the first few
  billion years of their lives.
  %
  Here we apply this knowledge to stars observed by the primary Kepler
  mission in order to derive rotation and lithium-based ages.
  %
  Rotation periods are drawn from previous Kepler analyses; lithium is
  measured from new and archival HIRES spectra.
  %
  We report rotation-based ages for \nuniqstarsantosrotgyroappl\ stars
  and \nplwgyroagenograzing\ planets for which we believe
  gyrochronology is valid;
  these rotation-based ages are $\approx$15-20$\times$ more precise
  than isochrone ages for our sample of FGK stars.
  %
  While the independent lithium and rotation-based ages agree
  for $\approx$90\% of stars, the timescales of lithium depletion are
  such that rotation is the more precise age diagnostic for the
  majority of our stellar sample.
  %
  Our resulting catalog of planetary ages includes
  \nplyounggyrotwosigma\ planets younger than 1\,Gyr at 2$\sigma$
  (\nplyounggyrothreesigma\ planets at 3$\sigma$).
  %
  For both planet-hosts, and for the broader sample of stars in the
  Kepler field, we find that the stellar age distribution decreases
  linearly from 3\,Gyr ago to present day, implying a linearly
  decreasing star formation rate.
  %
  While this parent stellar age distribution has implications for the
  star formation history of our Galaxy, it also bounds the relative
  balance of young and old planets detectable using both Kepler and
  also future wide-field transit surveys.
  %
\end{abstract}

@@

\subsubsection{Rotation vs.~Lithium Age Comparison}

We judged the consistency of the independent rotation and
lithium-based ages by comparing the point-estimates of the posteriors,
when available.
There are three cases of interest:

\begin{enumerate}
  \item Two-sided posteriors available for both $t_{\rm gyro}$ and
    $t_{\rm Li}$: then if their values were consistent within
    2$\sigma$, we judged them to be consistent; if they were
    consistent within 2-3$\sigma$, we suggest them to be ``maybe''
    consistent.  Otherwise, they are in disagreement.
\item If $t_{\rm Li}$ is a lower limit, we compare the 3$\sigma$
  $t_{\rm Li}$ lower limit with the $1\sigma$ upper limit from $t_{\rm
    gyro}$; if the two overlapped, we judge the age estimates to be
    consistent.
\item If $t_{\rm Li}$ provides a two-sided posterior, and no rotation
  is reported, we judge the age estimates to be inconsistent.
  This is because such constraints can only be provided for G and K
  stars $\lesssim$2\,Gyr old, and Kepler should have been sensitive to
  the rotation periods of such stars.
\end{enumerate}

In the entire sample, this yielded 
\allagesyesconsistent\ consistent cases,
\allagesmaybeconsistent\ ``maybe'' cases, and
\allagesnoconsistent\ inconsistent.
Restricting our attention to the $\lesssim$1\,Gyr stars, for which
lithium ages are more constraining,
the relevant numbers are
\minageltonegyryesconsistent, \minageltonegyrmaybeconsistent, and
\minageltonegyrnoconsistent.
Rephrased, in the overall sample, \fracconsistentallages\%
of stars have consistent rotation and lithium-based ages
(\fracpotentiallyconsistentallages\% potentially consistent).
In the sample of sub-gigayear stars,
\fracconsistentminageltonegyr\ of these ages are
consistent (\fracpotentiallyconsistentminageltonegyr\% potentially
consistent).
We feel it necessary to distinguish ``maybe'' consistent cases because
of lingering uncertainties in how the dispersion of the lithium data
evolve over the first gigayear.
In Appendix~\ref{app:inconsistent}, we discuss the
\minageltonegyrnoconsistent\ discrepant cases in detail.

@@

@@

\startlongtable
\begin{deluxetable*}{llllllllrrrrrrl}
	%\tabletypesize{\footnotesize}
	\tabletypesize{\scriptsize}
  \tablecaption{Ages of Kepler planets and planet candidates.  This
  version of the table is truncated to include the youngest fifty
  confirmed Kepler planets, and the twenty youngest candidates,
  sorted by the minimum of either the rotation or lithium-based
  age.  The
  full machine-readable table contains ages and age limits for
  \nnonfopkoissomeageinfo\ non-false positive KOIs with MES$>$10.  
  \label{tab:planets}}
	%\toprule
	%\midrule
	%\endhead
	\startdata
  KOI & Kepler &  $T_{\rm eff}$ & $P_{\rm rot}$ & EW$_{\rm Li}$ &
  $t_{\rm gyro}$ & $t_{\rm Li} $ & Consistent? & $R_{\rm p}$ & $P_{\rm orb}$ &
  $Q_{\rm planet}$ & $Q_{\rm gyro}$ &  Spec? & Comment \\
  -- &   -- & K & days & m\AA & Myr & Myr &  str &   Earths &    days &       int  & int & bool & -- \\
	\hline
	\input{table_subgyr_confirmed} %\\
%	\rule
	\input{table_subgyr_candidate}
	\enddata
	\tablecomments{``Confirmed'' planets appear before``candidate'' planets.
		The bit quality flags for the rotation-based ages,
		$Q_{\rm gyro}$, are described in Section~\ref{subsec:flags}.
		Concisely summarized, they are:
		Bit 0: $T_{\rm eff}/{\rm K} \in [ 3800-6200]$ ?
		Bit 1: $\log g$$<$4.2?
		Bit 2: $M_{\rm G}$<3.9?
		Bit 3: In KEBC?
		Bit 4: Large $d_{\rm xm,Kep-Gaia}$?
		Bit 5: Confused Kep-Gaia crossmatch?
		Bit 6: Gaia DR3 non-single star?
		Bit 7: RUWE$>$1.4?
		Bit 8: Crowded?
		Bit 9: Far from main-sequence?
		Bit 10: \citetalias{Santos_2021} CP/CB?
		As an example, Kepler-1627 has $Q_{\rm gyro}$ flagged with bit 10 and bit 7.
		The analogous planet quality bit-flag, $Q_{\rm planet}$, has the
		following meaning.
		Bit 0: Candidate not reliable?
		Bit 1: MES$<$10?
		Bit 2: Grazing?
		For a star to have a high likelihood of being ``reliable for gyrochronology'' we suggest $Q_{\rm star}$ bits
		zero to nine to not be raised, and for a planet to be ``reliabe'', we suggested $Q_{\rm planet}$ to be zero.
	}
\end{deluxetable*}

@@

\appendix
\section{Cases With Inconsistent Rotation and Lithium-Based Ages}
\label{app:inconsistent}

% from inconsitent_tli_tgyro cases:
Kepler-1644 b
Kepler-1699 b
 Kepler-786 b

these have finite Li ages, but no rotation age (perhaps b/c of the missing prots...)
Kepler-1943 b
 Kepler-101 b
 Kepler-108 b
 Kepler-108 c
  Kepler-63 b
Kepler-1312 b
Kepler-1312 c
   Kepler-1 b

K02680.01
K02639.01
